### 2014-07-20 docs/paper/text/problem.tex
added the detailed description of the ILP for the case of non-overlapping partitions

\paragraph{\textbf{Other useful equations}}

\noindent The following are unused, but may be needed for the partitioning
algorithm.

Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
attribute-timestamp pair $(a, t)$ to the frequency of queries that access
attribute $a$ for time $t$. More formally:
\begin{equation}
f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
\end{equation}

Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
attribute access for the time interval $T$ for attribute $a$ as follows:
\begin{equation}
f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
\end{equation}

>>>>

%\paragraph{\textbf{Other useful equations}}
%\noindent The following are unused, but may be needed for the partitioning
%algorithm.
%
%Let $f: \mathbb{R}\times A \rightarrow \mathbb{N}$ be a function that maps a
%attribute-timestamp pair $(a, t)$ to the frequency of queries that access
%attribute $a$ for time $t$. More formally:
%\begin{equation}
%f(a,t) = \sum_{q\in Q} \textbf{1}(a\in q.A \wedge t\in q.T)
%\end{equation}
%
%Let $T$ represent a time range $[T.s, T.e]$. We denote the frequency of
%attribute access for the time interval $T$ for attribute $a$ as follows:
%\begin{equation}
%f(a, T) = \int_{T.s}^{T.e} f(a, t)\cdot dt =  \sum_{q\in Q} |q.T \cap T| 
%\end{equation}

### 2014-07-20 docs/paper/text/solution.tex
added the detailed description of the ILP for the case of non-overlapping partitions

\section{Partitioning}
This section is still a work-in-progress. We are trying to find the algorithms.

@@

\paragraph*{Problem.}\emph{Find a true partitioning of attributes that minimizes
  the query I/O and bounds the storage cost by some upper limit.}

\subsubsection{Mixed Integer Linear Program Formulation}
$x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.\\
$y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.\\
$z_{a,p,q}$: $1$ if partition $p$ is used by query $q$, and attribute $a$ is in partition $p$, $0$ otherwise.\\
$u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.\\

Let $q(a)\equiv \mathbf{1}(a \in q.A)$.

% NOTE: In the simulation we will have query weights that will factor into the objective function
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}

Here we use the following LP construction. Assume we have two variables,
$\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$ and $\beta_1\geq 0$. We want to
implement the following constraint: 
$$
\beta_2 = 
\begin{cases}
0 & \beta_1 = 0 \\
1 & \beta_1 \geq 0 
\end{cases}
$$

This could be expressed as a linear constraint as follows,  where $K$ is a
large constant guaranteed to be larger  that $\beta_1$ for practical purposes:
\begin{eqnarray}
\hspace{2cm}&& \beta_1 - \beta_2 \geq 0\nonumber\\
\hspace{2cm}&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
\end{eqnarray}

In our ILP formulation, we use this construction twice. The first use is shown
in the second and third constrants, where $\beta_1=\sum_{a\in A} q(a)\cdot
x_{a,p}$ and $\beta_2=y_{p,q}$. The second use is shown in the fifth and 
sixth constaints, where $\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.

@@

\paragraph*{Approach.} TODO

@@

\subsection{Overlapping Attributes}

\paragraph*{Problem.}\emph{Find an overlapping partitioning of attributes that minimizes
  the query I/O and bounds the storage cost by some upper limit.}

\subsubsection{Mixed Integer Linear Program Formulation}
% NOTE: In the simulation we will have query weights that will factor into the objective function
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} > 0\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} \leq k\nonumber\\
\forall_{\{a,q\}\in A\times Q},
    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && z_{a,p,q} - (y_{p,q} + q(a)) \geq -1 \nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}

>>>>

\clearpage
\newpage
\section{Partitioning}
This section is still a work-in-progress. We are trying to find the algorithms.

@@

\paragraph*{Problem.} \emph{Find a true partitioning of attributes that
minimizes the query I/O and bounds the storage cost by some upper limit.}

\subsubsection{Integer Linear Program Formulation}
We present an ILP formuation of the problem. For this purpose, we define a 
number of binary ($0$ or $1$) variables: 
\begin{itemize}
\item $x_{a,p}$: $1$ if attribute $a$ is in partition $p$, $0$ otherwise.
\item $y_{p,q}$: $1$ if partition $p$ is used by query $q$, $0$ otherwise.
\item $z_{a,p,q}$: $1$ if partition $p$ is used by query $q$ and attribute $a$
is in partition $p$, $0$ otherwise.
\item $u_{p}$: $1$ if partition $p$ is assigned at least $1$ attribute, $0$ otherwise.
\end{itemize}

Each of these variables serve an important purpose:
\begin{itemize}
\item $x$s define the attribute to partition assignment.
\item $y$s help formulate the objective function's I/O contributions
that are on a per-partition basis.
\item $z$s help formulate the objective function's I/O contributions
that are on a per-attribute basis.
\item $u$s help formulate the storage overhead requirement.
\end{itemize}

We also define a helper notation for representing whether a variable is
accessed by a query or not: $q(a)\equiv \mathbf{1}(a \in q.A)$, where
$\mathbf{1}$ is the indicator function. We assume that there are a maximum of
$k$ partitions. $k$ can be taken as $|A|$, as there cannot be more partitions
than there are attributes.

We are now ready to state the ILP formulation. We start with the objective
function, that is the total query I/O cost:
\begin{eqnarray}
\sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot
y_{p,q}\nonumber\\ 
&+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q}\label{eq:no-obj}
\end{eqnarray}

In Eq.~\ref{eq:no-obj}, we simply sum for each query and each partiton to add
the I/O cost of reading in the structural information found in a sub-block, if
the partition is used by the query. We then sum over each attribute as well,
and add the I/O cost of reading in the attributes. Note that $z_{a,p,q}$ could
have been  replaced with $x_{a,p}\cdot y_{p,q}$, but that would make the
objective function non-linear.

Our first constraint is that each attribute gets assigned to one and only one
partition. Formally:
\begin{eqnarray}
\forall_{a\in A}, \sum_{p=1}^{k} x_{a,p} = 1
\end{eqnarray}

Our second constraint is that, if a query contains an attribute assigned to a
partition, then that partition is used by the query. In essence, we want to
state: $\forall_{\{p,q\}\in [1..k]\times Q}, y_{p,q} = \mathbf{1}(\sum_{a\in
A} q(a)\cdot x_{a,p}>0)$. In order to formulate this we use the following ILP 
construction:

Assume we have two variables, $\beta_1$ and $\beta_2$, where $\beta_2\in[0,1]$
and $\beta_1\geq 0$. We want to implement the following constraint: $\beta_2 =
\mathbf{1}(\beta_1 > 0)$.

This could be expressed as a linear constraint as follows, where $K$ is a
large constant guaranteed to be larger than $\beta_1$ for practical purposes:
\begin{eqnarray}
&& \beta_1 - \beta_2 \geq 0\nonumber\\
&& K\cdot\beta_2 - \beta_1 \geq 0\nonumber
\end{eqnarray}

We now apply this construction to our second constraint, where
$\beta_1=\sum_{a\in A} q(a)\cdot x_{a,p}$ and $\beta_2=y_{p,q}$. This results
in the following linear constraints:
\begin{eqnarray}
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 
\end{eqnarray}

Our third constraint is that, if an attribute $a$ is assigned to a partition
$p$, and partition $p$ is used by a query $q$, then the corresponding $z$
variable must be set to $1$. It must be $0$ in all other cases. That is, we
want: $\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
z_{a,p,q}=\mathbf{1}(x_{a,p} = y_{p,q} = 1)$. We express this as a linear 
constraint, as follows:
\begin{eqnarray}
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\label{eq:no-z}
\end{eqnarray}

In Eq.~\ref{eq:no-z}, when the $x$ and $y$ variable are both $1$, the  $z$
variable is simply forced to be $1$. Otherwise, the $z$ variable can be either
$0$ or $1$, but since the $z$ variables appear in the objective fucntion as
positive terms, the solver will set them to $0$, which is what we want.

Our fourth constraint is that, if a partition is non-empty, then its
corresponding $u$ variable must be set to $0$. It must be $0$ otherwise. In
other words,  we want $u_p = \mathbf{1}(\sum_{a\in A} x_{a,p}>0)$. This is
expressed as linear constraints, as follows:
\begin{eqnarray}
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \label{eq:no-u}
\end{eqnarray}

Eq.~\ref{eq:no-u} uses the same construction as the second constraint, where
$\beta_1=\sum_{a\in A} x_{a,p}$ and $\beta_2=u_p$.

Our fifth, and the last, constraint deals with the storage overhead. We want to
 make sure that the storage overhead does not go over $\alpha$. The storage
overhead depends on the number of partitions used. That means that the only 
ILP variables it depends on is the $u$s. In particular, the number of
partitions used is given by $\sum_{p=1}^{k} u_p$. This results in the
following linear constraint:
\begin{equation}
\sum_{p=1}^{k} u_p - 1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}}
\end{equation}

% NOTE: In the simulation we will have query weights that will factor into the objective function
\begin{figure}[!t]
\begin{mdframed}
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} = 1\nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} q(a)\cdot x_{a,p} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} q(a)\cdot x_{a,p}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q},
    && z_{a,p,q} - (x_{a,p} + y_{p,q}) \geq -1\nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}
\end{mdframed}
\caption{ILP formulation for the non-overlapping partitioning}
\label{fig:no-ilp}
\end{figure}

The final ILP formulation for the non-overlapping partitioning is given in
Figure~\ref{fig:no-ilp}.

@@

\paragraph*{Approach.$\,$}
TODO

@@

\clearpage
\newpage
\subsection{Overlapping Attributes}

\paragraph*{Problem.} \emph{Find an overlapping partitioning of attributes that minimizes
  the query I/O and bounds the storage cost by some upper limit.}

\subsubsection{Integer Linear Program Formulation}
% NOTE: In the simulation we will have query weights that will factor into the objective function
\begin{eqnarray}
\text{minimize}  
    \sum_{q\in Q} \sum_{p=1}^{k} && (16\cdot c_e(B) + 12\cdot c_n(B))\cdot y_{p,q}\nonumber\\
    &+& \sum_{a\in A} s(a)\cdot c_e(B)\cdot z_{a,p,q} \nonumber\\
\text{subject to}&&\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} > 0\nonumber\\
\forall_{a\in A}, 
    && \sum_{p=1}^{k} x_{a,p} \leq k\nonumber\\
\forall_{\{a,q\}\in A\times Q},
    &&  \sum_{p=1}^{k} z_{a,p,q} \geq q(a) \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  \sum_{a\in A} z_{a,p,q} - y_{p,q} \geq 0 \nonumber\\
\forall_{\{p,q\}\in [1..k]\times Q}, 
    &&  K\cdot y_{p,q} - \sum_{a\in A} z_{a,p,q}  \geq 0 \nonumber\\
\forall_{\{a,p,q\}\in A\times [1..k]\times Q}, 
    && z_{a,p,q} - (y_{p,q} + q(a)) \geq -1 \nonumber\\
\forall_{p\in[1..k]},
    && \sum_{a\in A} x_{a,p} - u_p \geq 0 \nonumber\\
\forall_{p\in[1..k]},
    && K\cdot u_p - \sum_{a\in A} x_{a,p} \geq 0 \nonumber\\    
&& \sum_{p=1}^{k} u_p -1 \leq \frac{\alpha}
  {1-\frac{c_e(B)\cdot \sum_{a\in A} s(a)}{s(B)}} \nonumber
\end{eqnarray}
